Simulate regression example

### 2022-07-10 notebook/_posts/2021-08-03-simulate-regression-example/simulate-regression-example.Rmd
Update regression example figure
Updated the code and figure to use the new terminology of 'abundance'
instead of 'density' and 'fold difference' instead of 'fold change'

## Actual densities

@@

```
log density = a_0 + a_1 * X + epsilon
```

@@

```{r}
sam <- tibble(.sample = colnames(y), x) %>%
  mutate(across(x, factor))
actual_dens <- phyloseq(
  otu_table(y, taxa_are_rows = TRUE) %>% transform_sample_counts(exp),
  sample_data(sam)
)
```

```{r}
actual_dens %>%
  as_tibble %>%
  ggplot(aes(y = .otu, x = .abundance, fill = x)) +
  scale_x_log10() +
  stat_dots()
```

## Taxonomic bias and measured densities

@@

The measured (i.e. estimated) proportions and densities are given by perturbing the actual densties by the efficiencies, and normalizing to proportions or to the original (correct) total.

```{r}
measured_prop <- actual_dens %>% perturb(efficiency)
measured_dens <- actual_dens %>% perturb(efficiency, norm = "keep")
```

@@

```{r mean-efficiency}
mean_eff <- actual_dens %>%
  transform_sample_counts(close_elts) %>%
  perturb(efficiency, norm = "none") %>%
  sample_sums %>%
  enframe(".sample", "mean_efficiency") %>%
  left_join(sam, by = ".sample") %>%
  mutate(log2_mean_efficiency = log2(mean_efficiency))
```

@@

## Estimated and actual densities

@@

dens <- bind_rows(
  Actual = actual_dens %>% as_tibble,
  Measured = measured_dens %>% as_tibble,
  .id = 'type'
) %>%
  rename(density = .abundance) %>%
  mutate(
    log2_density = log2(density),
    across(.otu, factor, levels = lvls)
  )
```

```{r}
p_species_all <- dens %>%
  ggplot(aes(x, log2_density, color = type)) +
  labs(y = "Log density ") +
  facet_grid(.otu~type, scales = "fixed") +
  geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
  # stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
  stat_summary(geom = "point") +
  stat_smooth(aes(x = as.integer(x)),
    method = "lm", size = 0.9, fill = 'grey', se = FALSE
  ) +
  theme(legend.position = "none")
p_species_all 
```

@@

```{r}
lm_fits <- dens %>%
  group_by(.otu, type) %>%
  nest %>%
  mutate(
    fit = map(data, ~lm(log2_density ~ 1 + x, .))
  )
lm_results <- lm_fits %>%
  mutate(
    fit = map(fit, broom::tidy, conf.int = TRUE)
  ) %>%
  select(-data) %>%
  unnest(fit) %>%
  ungroup %>%
  left_join(true_coeffs, by = c(".otu", "term"))

@@

p_coef_ci <- lm_results_slope %>% 
  ggplot(aes(y = .otu, x = estimate, color = type)) +
  labs(x = "Log fold change", y = NULL) +
  geom_vline(xintercept = 0, color = "grey") +
  geom_pointinterval(aes(xmin = conf.low, xmax = conf.high)) +
  theme(legend.position = 'top') +
  guides(color = guide_legend(reverse = TRUE)) +
  annotate(
    geom = 'segment', color = "darkgrey",
    arrow = grid::arrow(length = unit(0.1, "inches")), 
    x = start, xend = start - delta, 
    y = 10.5, yend = 10.5
  ) +
  coord_cartesian(clip = 'off')

@@

p_coef_ci_mean_eff <- lm_results_mean_eff %>% 
  filter(term == "x1") %>%
  ggplot(aes(y = "Mean efficiency", x = estimate)) +
  expand_limits(
    x = c(
      min(lm_results_slope$conf.low), 
      max(lm_results_slope$conf.high)
  )) +
  labs(x = "Log fold change", y = NULL) +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point() +
  annotate(
    geom = 'segment', color = "darkgrey",
    arrow = grid::arrow(length = unit(0.1, "inches")), 
    x = 0, xend = delta, 
    y = 1.0, yend = 1.0
  )

p_coef_dot <- lm_results_slope %>% 
  ggplot(aes(y = type, x = estimate, fill = type)) +
  labs(x = "Log fold change", y = "Type") +
  geom_vline(xintercept = 0, color = "grey") +
  stat_dots()

@@

# Set range of mean eff panel to match
rng <- p_species_focal$data %>%
  pull(log2_density) %>%
  {max(.) - min(.)}
m <- p_mean_eff$data %>% pull(mean_efficiency) %>% log2 %>% mean
p_mean_eff1 <- p_mean_eff + 
  expand_limits(y = c(m - rng/2, m + rng/2))
```

```{r main-figure, fig.dim = c(7, 7)}
(p_mean_eff1 + ggtitle("Mean efficiency")) +
  (p_coef_ci_mean_eff + ggtitle("LFC in mean efficiency")) + 
  (p_species_focal + ggtitle("Actual and measured densities\nof select species")) + 
  (p_coef_ci1 + ggtitle("Estimated LFCs of all species") +
    theme(legend.box.margin = margin(b = -15))
  ) + 
  plot_layout(ncol = 2, heights = c(0.2, 1)) +
  plot_annotation(tag_levels = 'A') &
  colorblindr::scale_color_OkabeIto() &
  colorblindr::scale_fill_OkabeIto() &
  theme(
    plot.title = element_text(face = "plain")
  )
```

@@

Can see that the standard error is generally increased in the measured density estimates except for Species 9, whose residual variation we expect to be correlated with the log mean efficiency and hence negatively correlated with the measurement error.

@@

```{r}
aug_mean_eff <- lm_fit_mean_eff %>% 
  broom::augment() %>% 
  add_column(.sample = mean_eff$.sample)
aug <- lm_fits %>%
  mutate(
    .sample = map(data, pull, '.sample'),
    augment = map(fit, broom::augment),
  ) %>%
  select(-data, -fit) %>%
  unnest(c(.sample, augment)) %>%
  ungroup %>%
  left_join(aug_mean_eff, by = c(".sample", "x"), suffix = c('.dens', '.mean_eff'))
```

```{r}
cov_actual_with_mean_eff <- aug %>%
  filter(type == 'Actual') %>%
  with_groups(.otu, summarize,
    cov = cov(.resid.dens, .resid.mean_eff)
  ) %>%
  print
```

>>>>

## Actual abundances

@@

```
log abundance = a_0 + a_1 * X + epsilon
```

@@

```{r}
sam <- tibble(.sample = colnames(y), x) %>%
  mutate(across(x, factor))
actual_abun <- phyloseq(
  otu_table(y, taxa_are_rows = TRUE) %>% transform_sample_counts(exp),
  sample_data(sam)
)
```

```{r}
actual_abun %>%
  as_tibble %>%
  ggplot(aes(y = .otu, x = .abundance, fill = x)) +
  scale_x_log10() +
  stat_dots()
```

## Taxonomic bias and measured abundances

@@

The measured (i.e. estimated) proportions and abundances are given by perturbing the actual abundances by the efficiencies, and normalizing to proportions or to the original (correct) total.

```{r}
measured_prop <- actual_abun %>% perturb(efficiency)
measured_abun <- actual_abun %>% perturb(efficiency, norm = "keep")
```

@@

```{r mean-efficiency}
mean_eff <- actual_abun %>%
  transform_sample_counts(close_elts) %>%
  perturb(efficiency, norm = "none") %>%
  sample_sums %>%
  enframe(".sample", "mean_efficiency") %>%
  left_join(sam, by = ".sample") %>%
  mutate(log2_mean_efficiency = log2(mean_efficiency))
```

@@

## Estimated and actual abundances

@@

abun <- bind_rows(
  Actual = actual_abun %>% as_tibble,
  Measured = measured_abun %>% as_tibble,
  .id = 'type'
) %>%
  rename(abundance = .abundance) %>%
  mutate(
    log2_abundance = log2(abundance),
    across(.otu, factor, levels = lvls)
  )
```

```{r}
p_species_all <- abun %>%
  ggplot(aes(x, log2_abundance, color = type)) +
  labs(y = "Log abundance ") +
  facet_grid(.otu~type, scales = "fixed") +
  geom_quasirandom(alpha = 0.3, groupOnX = TRUE) +
  # stat_summary(fun.data = "mean_cl_boot", geom = "pointrange") +
  stat_summary(geom = "point") +
  stat_smooth(aes(x = as.integer(x)),
    method = "lm", size = 0.9, fill = 'grey', se = FALSE
  ) +
  theme(legend.position = "none")
p_species_all 
```

@@

```{r}
lm_fits <- abun %>%
  group_by(.otu, type) %>%
  nest %>%
  mutate(
    fit = map(data, ~lm(log2_abundance ~ 1 + x, .))
  )
lm_results <- lm_fits %>%
  mutate(
    fit = map(fit, broom::tidy, conf.int = TRUE)
  ) %>%
  select(-data) %>%
  unnest(fit) %>%
  ungroup %>%
  left_join(true_coeffs, by = c(".otu", "term"))

@@

p_coef_ci <- lm_results_slope %>% 
  ggplot(aes(y = .otu, x = estimate, color = type)) +
  labs(x = "Log fold difference", y = NULL) +
  geom_vline(xintercept = 0, color = "grey") +
  geom_pointinterval(aes(xmin = conf.low, xmax = conf.high)) +
  theme(legend.position = 'top') +
  guides(color = guide_legend(reverse = TRUE)) +
  annotate(
    geom = 'segment', color = "darkgrey",
    arrow = grid::arrow(length = unit(0.1, "inches")), 
    x = start, xend = start - delta, 
    y = 10.5, yend = 10.5
  ) +
  coord_cartesian(clip = 'off')

@@

p_coef_ci_mean_eff <- lm_results_mean_eff %>% 
  filter(term == "x1") %>%
  ggplot(aes(y = "Mean efficiency", x = estimate)) +
  expand_limits(
    x = c(
      min(lm_results_slope$conf.low), 
      max(lm_results_slope$conf.high)
  )) +
  labs(x = "Log fold difference", y = NULL) +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point() +
  annotate(
    geom = 'segment', color = "darkgrey",
    arrow = grid::arrow(length = unit(0.1, "inches")), 
    x = 0, xend = delta, 
    y = 1.0, yend = 1.0
  )

p_coef_dot <- lm_results_slope %>% 
  ggplot(aes(y = type, x = estimate, fill = type)) +
  labs(x = "Log fold difference", y = "Type") +
  geom_vline(xintercept = 0, color = "grey") +
  stat_dots()

@@

# Set range of mean eff panel to match
rng <- p_species_focal$data %>%
  pull(log2_abundance) %>%
  {max(.) - min(.)}
m <- p_mean_eff$data %>% pull(mean_efficiency) %>% log2 %>% mean
p_mean_eff1 <- p_mean_eff + 
  expand_limits(y = c(m - rng/2, m + rng/2))
```

```{r main-figure, fig.dim = c(7, 7)}
(p_mean_eff1 + ggtitle("Mean efficiency")) +
  (p_coef_ci_mean_eff + ggtitle("LFD in mean efficiency")) + 
  (p_species_focal + ggtitle("Actual and measured\nabundances of select species")) + 
  (p_coef_ci1 + ggtitle("Estimated LFDs of all species") +
    theme(legend.box.margin = margin(b = -15))
  ) + 
  plot_layout(ncol = 2, heights = c(0.2, 1)) +
  plot_annotation(tag_levels = 'A') &
  colorblindr::scale_color_OkabeIto() &
  colorblindr::scale_fill_OkabeIto() &
  theme(
    plot.title = element_text(face = "plain")
  )
```

@@

Can see that the standard error is generally increased in the measured abundance estimates except for Species 9, whose residual variation we expect to be correlated with the log mean efficiency and hence negatively correlated with the measurement error.

@@

```{r}
aug_mean_eff <- lm_fit_mean_eff %>% 
  broom::augment() %>% 
  add_column(.sample = mean_eff$.sample)
aug <- lm_fits %>%
  mutate(
    .sample = map(data, pull, '.sample'),
    augment = map(fit, broom::augment),
  ) %>%
  select(-data, -fit) %>%
  unnest(c(.sample, augment)) %>%
  ungroup %>%
  left_join(aug_mean_eff, by = c(".sample", "x"), suffix = c('.abun', '.mean_eff'))
```

```{r}
cov_actual_with_mean_eff <- aug %>%
  filter(type == 'Actual') %>%
  with_groups(.otu, summarize,
    cov = cov(.resid.abun, .resid.mean_eff)
  ) %>%
  print
```
